---
title: "Untitled"
author: "Kathleen DeBrota"
date: "March 5, 2019"
output: pdf_document
---
#Midterm exam questions and answers.
## contains data description, wrangling, dplyr/tidying, importing, finding missing values, etc

```{r}
#set the working directory: 
setwd("C:/Users/kdebr/OneDrive/Applied Data Science Class") 

#import the csv file:
epa<- read.table("US EPA data 2017.csv", header=TRUE, sep = ",")

#check the datatypes:
#str(epa)

glimpse(epa)

```
Question 1: how many columns have missing values?

```{r}
#replace empty datapoints with NA
#is.na(epa)

library(dplyr)
#Note: the method below yielded '7', but this is not an answer choice...there are at least 8...
#epa %>% select_if(function(x) any(is.na(x))) %>%
#  summarize_each(funs(sum(is.na(.))))

#epa %>% select_if(function(epa)any(is.na(epa))) %>%
 # summarize(funs(sum(is.na(.))))

#second attempt
colSums(is.na(epa))
```
Once again, this method shows there are 7 columns that contain >0 missing values. Now attempting to import the data as an excel file instead
```{r}
library(readxl)
epa2 <- read_excel("US EPA data 2017.xlsx")
head(epa2)
```
Now trying to find NA's:
```{r}
colSums(is.na(epa2))
```
Skipping to question 18:
```{r}
iris <- iris
#iris
iris %>% filter(Petal.Length > 4.5) %>% group_by(Species) %>% summarize(mean_sepallength=mean(Sepal.Length), sd_sepalwidth=sd(Sepal.Width), max_petallength=max(Petal.Length))
```
Question 19: using diamonds dataset...
```{r}
library(ggplot2)
diamonds <- diamonds
sorted <- table(diamonds$color)
sorted
```
Question 20:
```{r}
library(nycflights13)
flights <- flights
#head(flights)
flights %>% filter(month == 1 | month == 6)

#and to check answer...
flights %>% filter(month==1)
flights %>% filter(month==6)
#yep, answer is correct.
```
Question 17:find 25 percentile of the eruptions column
```{r}
faithful <- faithful
quantile(faithful$eruptions, 0.25)
```
Question 12:
```{r}
library(tidyverse)
head(who)
```
Question 13:
```{r}
library(ggplot2)
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy), size=3, shape=24, color="black", fill = 'red')
```

Question 7:
```{r}
mtcars <-mtcars
#head(mtcars)
library(matrixStats)
mtcars %>% summarize(
  mean_mpg=mean(mtcars$mpg), 
  trimmed_mpg=mean(mtcars$mpg, trim=0.1),
  weighted_mean=weighted.mean(mtcars$mpg, mtcars$wt),
  median_mpg=median(mtcars$mpg)
)

weighted.median(mtcars$mpg, mtcars$wt)

#  sd_mpg=sd(mtcars$mpg), 
 # IQR_mpg=IQR(mtcars$mpg), 
  #mad_mpg=mad(mtcars$mpg)
  #)
```
Question 4:
```{r}
mtcars %>% summarize(sd_mpg=sd(mtcars$mpg), IQR_mpg=IQR(mtcars$mpg), mad_mpg=mad(mtcars$mpg))
mtcars
```
Question 2:
```{r}


#mtcars$cyl <- NULL
#mtcars$gear <-NULL
#mtcars$carb <- NULL
#mtcars$am <- NULL
#mtcars$vs <- NULL
head(mtcars)

library(corrplot)

matrix <- cor(mtcars)
matrix


```
mpg-wt = -.8676
disp-wt = 0.8879
mpg-disp = -0.8475
disp-hp = 0.7909
