---
title: "For quiz 6 applied datasci"
author: "Kathleen DeBrota"
date: "February 13, 2019"
output: pdf_document
---
# Basic stats, dplyr commands, data description

Calculating the trimmed mean: drop a fixed number (p) of sorted values at each end, and then take an average of the remaining values. 
This metric eliminates the influence of the most extreme values at either end. 
```{r}
library(readr)
library(dplyr)
setwd("C:/Users/kdebr/OneDrive/Applied Data Science Class")
#Note: this is how you set the current working directory. MUST use forward slashes!
incomedata <- read_csv("income.csv")
#str(incomedata)
mean(incomedata$'M_weekly', na.rm=TRUE)
median(incomedata$'M_weekly', na.rm=TRUE)

#calculating the trimmed mean: ordering the data
mean(incomedata[['M_weekly']], trim=0.1, na.rm=TRUE)

```
Calculating standard deviation, IQR, and median absolute standard deviation (MAD) for female workers. 
```{r}
#Interquartile range:
IQR(incomedata$'F_workers', na.rm=TRUE)
#standard deviation:
sd(incomedata$'F_workers', na.rm=TRUE)
#median absolute standard deviation from the median (MAD):
# this metric is a robust estimate of variability. It's not influenced by extreme values, and it is possible to also compute a trimmed standard deviation analogous to the trimmed mean. R has a built infunction for the MAD:
mad(incomedata$'F_workers', na.rm=TRUE)

```
### Using dplyr to mutate and make new columns
#### using 'glimpse'
Adding a new column named 'diff', the difference between male and female weekly income.
```{r}
incomedata <- mutate(incomedata, diff =incomedata$'M_weekly' - incomedata$'F_weekly')
#Peek at the dataset to see if the addition worked
incomedata %>% glimpse()
```
Calculating the standard deviation of diff column.
```{r}
sd_diff <- sd(incomedata$'diff', na.rm=TRUE)
round(sd_diff)
```
importing optdigits.tra; first turned into a .csv file in Excel and delimited comma-separated columns.
```{r}
optdigits <- read_csv("optdigits.csv")
shape(optdigits)
nrow(optdigits)
ncol(optdigits)
```

