---
title: "clustering in R"
author: "Kathleen DeBrota"
date: "April 17, 2019"
output: pdf_document
---

# Clustering 
## Following https://bookdown.org/rdpeng/exdata/k-means-clustering.html

Clustering is a way to partition data into groups which are refined further each time the algorithm is run. It picks random centroids ('k' of them), and assigns each point to its closest centroid. Then it recalculates centroid positions based on the new groups it found, and repeats until the centroids are no longer moving (or moving very little)

```{r}
set.seed(1234)
x<- rnorm(12, mean=rep(1:3, each=4), sd=0.2)
y<- rnorm(12, mean=rep(c(1,2,1), each=4), sd=0.2)
plot(x,y, col='blue', pch=19, cex=2)
text(x + 0.05, y + 0.05, labels=as.character(1:12))
#the above is just a simulated dataset. 

```
This practice starts by assuming there are 3 clusters. 
```{r}
#Therefore k = 3 here
dataFrame<-data.frame(x,y) #setting up the data into a df
kmeansObj <- kmeans(dataFrame, centers=3) #tell it you want 3 clusters
names(kmeansObj)
```
These names are the names of the different functions in kmeans()
```{r}
#Check which object was put into which cluster...
kmeansObj$cluster

#Find kmeans solution:
dataMatrix<- as.matrix(dataFrame)[sample(1:12),]
kmeansObj <- kmeans(dataMatrix, centers=3) #I guess you have to put it into a matrix first

#Plot the clusters as a HEAT MAP:
par(mfrow=c(1,2))
image(t(dataMatrix)[, nrow(dataMatrix):1], yaxt='n', main='Original data')
image(t(dataMatrix)[, order(kmeansObj$cluster)], yaxt='n', main='Clustered data')
```



