---
title: "Practice week 7 dataviz"
author: "Kathleen DeBrota"
date: "February 19, 2019"
output: pdf_document
---

practice for week 7 of applied data science course: data visualizations using ggplot2. 
```{r}
library(tidyverse)
```
We are trying to answer the question "Do cars with big engines use more fuel than cars with small engines?" Try to be precise, try to find a relationship between engine size and fuel efficiency if there is one. What does that relationship look like? Positive, negative, linear/nonlinear?
```{r}
#Load the mpg dataframe from ggplot2.
mpg
```
Plot mpg with displ on x axis andhwy on y-axis. hwy is the highway fuel efficiency and disp is the engine size. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x=displ, y=hwy))
```
Negative relationship between engine size and fuel efficiency: the larger the engine, the lower the efficiency, aka it uses more fuel to go the same distance.
This so far confirms our hypothesis.
The geom_point() function is the part that adds the points. 
In ggplot2, each function has a 'mapping' argument, which defines how you will map the variables onto the graph, aka how to translate the data into visualizations. map argument is always paired with aes() and x and y arguments of aes() specify which variables to map to which axes.
```{r}
#if we want a reusable template for making graphs
# ggplot(data = <DATA>) + 
# <GEOM_FUNCTION> (mapping= aes(<MAPPINGS>))
```
Practice exercises...
```{r}
#ggplot(data=mpg) #this gives an empty graph. there is no map argument.
mpg #there are 234 rows and 11 columns in the dataset.

#scatterplot of hwy vs cyl:
ggplot(data=mpg) +
  geom_point(mapping=aes(x=cyl, y=hwy))
#this scatterplot shows that the more cylinders an engine has, the lower its fuel efficiency -there is a negative trend, but a car has a fixed number of cylinders hence the grouped datapoints.
```
Are some of the largest-engine-yet-highest-fuel-efficiency cars 'outliers' or why are they so apparently separate from the rest of the data? Are they hybrids? We can check by looking at the 'class' column for each car. We want to add another variable to a two-dimensional scatterplot by mapping it to an aesthetic, = visual property of the objects in the plot. 
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy, color=class))
```
Now we see each different class of car in another color. This tells us that the two-seater cars are the group we were looking at, which are sportscars, which have a lot of engine power (large) but small body size (lightweight) so they are more fuel efficient per engine size than other cars. We can map class to 'size' instead:
```{r}
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy, size=class))
```
Or we can do this - the alpha aesthetic is transparency, or shape aesthetic which is the shape of points...
```{r}
#on the left:
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy, alpha=class))
#on the right:
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy, shape=class))
#noticing that only 6 shapes can be used at once, so the 7th class has been removed from this visualization
```
Setting aesthetic properties manually:
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy), color="purple")
#other color words that work: blue, red, green, cyan, pink, black, grey, navy, yellow, orange, magenta, turquoise,
```
NOTE make sure the manual aesthetic goes OUTSIDE of the aes() argument. colors must be character strings. size of points must be in mm. shape of a point as a number, as in a specific figure which gives 20 different shapes/colored shapes from squares/circles/X's, stars, etc.
```{r}
?geom_point
##ggplot(data=mpg)+
 # geom_point(mapping=aes(x=displ, y=hwy, stroke=class))
```
NOTE the + in the ggplot graphics lines has to come at the END of the line, NOT the start. It cannot be on the next level below, must be on the upper level above the next ggplot line.

FACETS: This is especially useful in categorical variables. Split plot into facets, which are subplots that each display one subset of the data. To facet by one single variable, use facet_wrap(). First argument is a formula, wihch you make with ~ followed by a variable name. The variable you pass to facet_wrap should be discrete (not continuous).
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  facet_wrap(~ class, nrow=2)
```

This gives one plot per 'facet', aka, 'class', in our dataset. 
Can facet by a combination of two variables: facet_grid(), with formula having two variable names separated by ~.
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  facet_grid(drv ~ cyl)
```
If you want to NOT facet in the rows or columns dimension, use a . instead of a variable name... as in... + facet_grid(.~cyl)
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  facet_grid(. ~ cyl)
#this chart faceted on the x axis only...so it shows engine size on x, with gas mileage on y, but separated into groups based on cylinder number (4, 5, 6, or 8).
```
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))+
  facet_grid(drv ~ .)
#this chart separated the opposite way, so it shows the drv variable as the facet, but with the same two variables on x and y axes.
```
GEOMETRIC OBJECTS
When two plots have the same x and y variables, and describe same data but aren't identical, they have different GEOMS. This is a geometrical object that a plot uses to represent data. Bar charts use bar geoms, etc. Change the geom function to change plot type.
```{r}
ggplot(data=mpg)+
  geom_smooth(mapping=aes(x=displ, y=hwy))
```

```{r}
ggplot(data=mpg)+
  geom_smooth(mapping=aes(x=displ, y=hwy, linetype = drv))
#this variation takes three separate lines based on their drv value (drivetrain). One line shows all points with a '4' value, another line shows all those with a 'f' value for drv, etc. 
```
```{r}
#OVERLAYING points and smooth lines
ggplot(data=mpg)+
  geom_smooth(mapping=aes(x=displ, y=hwy, linetype=drv))+
  geom_point(mapping=aes(x=displ, y=hwy, group = drv))
```
```{r}
#all of the above but group by color = class
ggplot(data=mpg, mapping=aes(x=displ, y=hwy))+
  geom_point(mapping=aes(color=class))+
  geom_smooth()
  
  
```
Using a filter at the same time:
```{r}
ggplot(data=mpg, mapping=aes(x=displ, y=hwy))+
  geom_point(mapping=aes(color=class))+
  geom_smooth(data=filter(mpg, class == 'subcompact'), se=FALSE)
#this ONLY shows the subcompact cars category in the 'smooth line' portion while keeping the rest of the categories as data points.
```
Bar charts:
This data set is showing 'cut' which is a variable from dataset 'diamonds'
```{r}
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut))
#NOTE new data set, = 'diamonds'
```
```{r}
ggplot(data=diamonds) +
  stat_count(mapping=aes(x=cut))
#Geoms and 'stats' are interchangeable in general. 
```
Might want to override default mapping from variables to aesthetics. for example you might want a bar chart of proportion, not count:
```{r}
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut, y= ..prop.., group=1))
#computed variables - look for help section with this title.
```
```{r}
ggplot(data=diamonds)+
  stat_summary(
    mapping=aes(x=cut, y=depth),
    fun.ymin=min,
    fun.ymax=max,
    fun.y=median
    )
```
Colors and position adjustments
```{r}
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut, color=cut))
#above is left graph
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut, fill=cut))
#above is right graph, yay colors! Note we use 'cut' for both the x and the fill variable...
```
```{r}
#adding another variable, cut and clarity together:
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut, fill=clarity))
#auto-stacked colors. this is a POSITION ADJUSTMENT 
```
```{r}
#other position adjustments...identity, dodge, and fill
ggplot(data=diamonds, mapping=aes(x=cut, fill=clarity)) +
  geom_bar(alpha=1/5, position="identity")
#another:
ggplot(data=diamonds, mapping=aes(x=cut, color=clarity)) +
  geom_bar(fill= NA, position="identity")
```
Position = "fill" makes every set of stacked bars the same height - good for proportions across groups. 
```{r}
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut, fill=clarity), position='fill')
```

```{r}
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut, fill=clarity), position='dodge')
#this way the overlapping objects are directly beside each other, not atop one another as in a stacked graph. this makes it easier to compare the individual values.
```
"jitter" in a scatterplot makes it so that there is a small amount of random noise to each point which spreads them out a little bit so there is less overlap. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")

```
MESSING WITH COORDINATE SYSTEMS:
```{r}
#these two plots are the same but with flipped axes. Useful for long labels. If they won't fit on one axis without overlapping...try flipping entire graph.
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
#above is left
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
#above is right graph
```
MAPS:
```{r}
#NOTE package is not installed, but this will fit the coordinates correctly with the right aspect ratio with geo-mapped data.
#nz <- map_data("nz")

#ggplot(nz, aes(long, lat, group = group)) +
##  geom_polygon(fill = "white", colour = "black")

#ggplot(nz, aes(long, lat, group = group)) +
 # geom_polygon(fill = "white", colour = "black") +
  #coord_quickmap()
```
OTHERS:
```{r}
bar <- ggplot(data = diamonds) + 
  geom_bar(
    mapping = aes(x = cut, fill = cut), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

